Can Symmetric Cryptography Be Liberated from the Von Neumann Style?

### 2024-10-16 paper.tex
Manage national pride
Clarifications

Moreover, ciphers are now, by and large, designed to be run
efficiently on software. The era of hardware-based cryptography is
long gone, even for embedded
platforms~\citep{buchanan:nist-lightweight-crypto}. To achieve
high-throughput, ciphers are typically designed so as to admit a
bitsliced implementation. Bitslicing enables a form of parallel
execution \emph{within} machine words. We thus gain access to a form
of CPU-level ``scale out'' to larger machine words, in particular
through single-instruction multiple-data (SIMD) instructions. Doubling
the size of machine words yields nearly twice the throughput, without
any effort. Evidence tends to suggest that cryptographers obtain
bitsliced designs through sheer intellectual might. It is in fact the
sole purpose of the traditional ``Section 4'' scene to exhibit a
bitsliced witness in all the gory details.

@@

We believe that there is also a broader take-away for an audience of
stubbornly French functional programmers. The present work surfs on
the wake of the Squiggol school of programming~\citep{bird:squiggol}.
Because our government-approved programming language does not yet
support ad-hoc polymorphism~\citep{white:modular-implicit}, the French
community may be missing out on some interesting programming patterns.
The present article can thus be read as a case-study in
``functor-oriented programming''~\citep{roconnor:functor-oriented},
where the notion of functor comes from category theory (recalled in
Section~\ref{sec:spec} and unrelated to the notion used in ML module
systems). In effect, we forbid ourselves from programming over a
concrete data structure: instead, we rely on (categorical) functors,
peppered with some more structure, so as to 1. unlock algebraic
manipulation during compilation and 2. exploit parametricity for
reasoning.

@@

We nonetheless could not resist the temptation of developing the
semantics in the Coq theorem prover, as executable programs. After
all, we are interested in combinational circuits: if there has ever
been a time where we do not have to worry about termination, this is
it! So we have implemented cryptographic primitive in pure Gallina.
This is nothing to worry about: after all, more ambitious colleagues
went from there down to deeply-embedded assembly
code~\citep{chlipala:coq-crypto}. Besides the ability to prove the
correctness and, even more usefully, test our code, this also enables
us to easily prototype a fixslicing compiler, using Coq's
\texttt{autorewrite} tactics to emulate a transformation-based
simplification engine~\citep{peyton-jones:haskell-optim} (as part of
an imaginary optimizing compiler).

>>>>

Moreover, ciphers are now, by and large, designed to be run
efficiently on software. The era of hardware-based cryptography is
long gone, even for embedded
platforms~\citep{buchanan:nist-lightweight-crypto}. To achieve
high-throughput, ciphers are typically designed so as to admit a
bitsliced implementation. Bitslicing enables a form of parallel
execution \emph{within} machine words, relying sometimes explicitly on
the availability of vectorized instructions~\citep{beaulieu:speck,
  bernstein:chacha}. We thus gain access to a form of CPU-level
``scale out'' to larger machine words, in particular through
single-instruction multiple-data (SIMD) instructions. Doubling the
size of machine words yields nearly twice the throughput, without any
effort. Evidence tends to suggest that cryptographers obtain bitsliced
designs through sheer intellectual might. It is in fact the sole
purpose of the traditional ``Section 4'' scene to exhibit a bitsliced
witness in all the gory details.

@@

We believe that there is also a broader take-away for an audience of
French functional programmers. The present work surfs on
the wake of the Squiggol school of programming~\citep{bird:squiggol}.
Because our government-approved programming language does not yet
support ad-hoc polymorphism~\citep{white:modular-implicit}, the French
community may be missing out on some interesting programming patterns.
The present article can thus be read as a case-study in
``functor-oriented programming''~\citep{roconnor:functor-oriented},
where the notion of functor comes from category theory (recalled in
Section~\ref{sec:spec} and unrelated to the notion used in ML module
systems). In effect, we forbid ourselves from programming over a
concrete data structure: instead, we rely on (categorical) functors,
peppered with some more structure, so as to 1. unlock algebraic
manipulation during compilation and 2. fully exploit parametricity for
reasoning.

@@

We nonetheless could not resist the temptation of developing the
semantics in the Coq theorem prover, as executable programs. After
all, we are interested in combinational circuits: if there has ever
been a time where we do not have to worry about termination, this is
it! So we have implemented cryptographic primitive in pure Gallina,
following the lead of some of our colleagues who went even further,
down to deeply-embedded assembly code~\citep{chlipala:coq-crypto}.
Besides the ability to prove the correctness and, even more usefully,
test our code, this also enables us to easily prototype a fixslicing
compiler, using Coq's \texttt{autorewrite} tactics to emulate a
transformation-based simplification
engine~\citep{peyton-jones:haskell-optim} (as part of a fictional
optimizing compiler).
